Stochastic Wind Power Forecasting

### 2020-05-30 LaTeX_Code/Paper/paper_1line_Renzo.tex
Corrected paper (added subsections 5.4.1, 5.)

\usepackage[backend=biber, style=authoryear, citestyle=authoryear, autocite=inline]{biblatex} % package for bibliography 
%\DeclareLanguageMapping{english}{english-apa}

@@

\newcommand{\add}{\textcolor{red}{[insert]}}

\setcounter{page}{1}
\sloppy
\begin{document}

@@

Previous attempt by (\cite{mozuma}) considered stochastic wind power forecast models based on stochastic differential equations. Here, we propose an improved model featuring time derivative tracking of the forecast, time-dependent mean reversion, modified diffusion and non-Gaussian approximations. We apply the model to Uruguayan wind power forecasts together with historical wind power production data pertaining to the year 2019.

{\color{red} Change this paragraph. Discuss with the global dataset, the dataset after removing days when curtailing has been detected, the data set without curtailing and with Lamperti transform.}
We have available a year long data set from Uruguay based on 363 observation paths, each of which is 24-hours long with observations recorded every 10 min. In total, it is a data set of approximately fifty thousand data points recorded in 2019. See Figure (\ref{fig:sample_data}). Data are normalized with respect to the maximum power capacity of wind power production in Uruguay, which is $\SI{1474}{\mega\watt}$. However, sometimes the real production is artificially modifies due to curtailing. This effect becomes clear when we inspect for different power production levels. We split the available data into low, medium and high power range. See Figure (\ref{fig:data_curtailing}).\\
We identify and remove the days with curtailing, removing a total of 108 days. From the resulting 255 days, we use 127 days to train the system and 128 to test it. Also, to avoid correlation between days, we intercalate the days we use for training and testing ({\color{red} Show that 24h is enough to ensure independence}).

In this paper we present the phenomenological underlying model in Section \ref{Section_2} and describe the physical constraints in Section \ref{Section_3} and how these constrains can be met. Then, in Section \ref{Section_4}, we will introduce an alternative formulation of the model in Lamperti space. In Section \ref{Section_5}, we show our parameter estimation procedure and its results in Section \ref{Section_6}. We compare alternative models in Section \ref{Section_7} and different forecast providers in Section \ref{Section_8}.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/5.eps}
\includegraphics[width=0.45\textwidth]{plots/245.eps}\\
\includegraphics[width=0.45\textwidth]{plots/661.eps}
\includegraphics[width=0.45\textwidth]{plots/805.eps}
\caption{Four samples from the Uruguayan data of 2019. Each sample compromises of two 24-hour paths. In blue is an hourly wind power production forecast. In orange is the actual wind power production recorded in 10 minute intervals.}
  \label{fig:sample_data}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_6.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_6.eps}
\caption{We see that forecast errors exhibit skewness, which is exaggerated as a consequence of curtailing in the production. Low power is when produced power is in $[0,0.3]$, mid-power is when it is in $(0.3,0.6]$, and high power when it is in $(0.6,1]$. We have a total of 363 days to use.}
  \label{fig:data_curtailing}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP.eps}
\includegraphics[width=0.45\textwidth]{plots/MP.eps}\\
\includegraphics[width=0.45\textwidth]{plots/HP.eps}
\includegraphics[width=0.45\textwidth]{plots/AP.eps}
\caption{We observe that skewness has been reduced after removing the days with curtailing. We have a total of 255 days with no curtailing.}
  \label{fig:data_after_clean}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_t.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_t.eps}\\
\includegraphics[width=0.45\textwidth]{plots/HP_t.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_t.eps}
\caption{Transitions in error between forecast and real production. We can observe that they are not totally centered around zero.}
\label{fig:error_transitions}
\end{figure}

@@

%---BEGIN SECTION 2---
\section{Phenomenological  Model} \label{Section_2}

In this section, once we analyzed the available dataset, we consider a type of phenomenological model for the normalized wind power generation forecasts which is given, in its most general form, as a stochastic process $X = \{X_t, \, t \in [0,T] \}$  defined by the following stochastic differential equation (SDE):

\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dX_t \!\!\!&=  a(X_t; p_t, \dot{p}_t,\bm{\theta}) \,dt + b (X_t; p_t, \dot{p}_t, \bm{\theta} ) \,dW_t\,, \quad t \in [0,T]  \\
     X_0  \!\!\!&=  x_0 \in [0,1],
  \end{array}\right. \label{eq:main}
\end{equation}

@@

\subsection{Physical Constraints} \label{Sec: Physical_Constraints}

@@

Remark: Observe that a mean-reverting model without derivative tracking shows a delayed path behavior. For instance, consider the diffusion model (\ref{eq:main}) with $a(X_t; p_t, \bm{\theta}) = - \theta_0 (X_t - p_t)\,, \theta_0 > 0$. In this case, given  $ \mathbb{E} X_0 = p_0$, the diffusion has mean $\mathbb{E} X_t = p_t - e^{- \theta_0 t } \int_0^t \dot{p}_s  e^{\theta_0 s} ds$. The next Figure (\ref{fig:derivative_tracking}) illustrates paths of two diffusions with and without derivative tracking.

\begin{figure}[h]
  %\includegraphics[width=26mm,scale=1]{}
  \caption{{\color{red} Add Figure showing the difference with and without derivative tracking}}
  \label{fig:derivative_tracking}
\end{figure}

@@

However, a distinctive feature in our proposed model 
\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dX_t \!\!\!&= (\dot{p}_t  - \theta_t (X_t - p_t) ) dt +\sqrt{2 \alpha \theta_0 X_t (1 - X_t)} dW_t\,, \quad t \in [0,T]  \\
   X_0  \!\!\!&=  x_0\in [0,1] \,,
 \end{array}\right.  \label{ourmodel}
\end{equation}

is that the drift term contains the time-varying parameter $\theta_t$, rendering the solution $X$ to (\ref{ourmodel}) a non-stationary and time-inhomogeneous process. To ensure that the process $X$ be the unique strong solution to (\ref{ourmodel}) for all $t \in [0,T]$ with state space $[0,1]$ a.s., the mean-reversion time-varying parameter must satisfy the following condition:
%has to be selected according to the following rule. Observe that the time derivative term $\dot{p}_t $ is not controlled to maintain that $X_t$ stays a.s. inside the range $[0,1]$. In other words, the zero drift line defined by $a(x; p_t,\bm{\theta}) =0$, which is an attractor, must be contained inside the range $[0,1]$. Thus, we must have that
%\begin{equation}
%\frac{- |\dot{p_t}|}{p_t} \leq \theta_t \leq \frac{|\dot{p}_t|}{1- p_t}
%\end{equation}
%which is satisfied  by choosing a time-dependent  $\theta_t$ as follows,
%\begin{equation}
%\theta_t = \max \left( \theta_0 \ , \ \frac{|\dot{p}_t|}{\min (p_t, 1-p_t)}  \right ),  \quad \theta_0 >0\,. \label{theta_t}
%\end{equation}
\begin{equation}\label{Assumption:3}
\theta_t\geq \max\left(\frac{\alpha\theta_0+\dot p_t}{1-p_t},\frac{\alpha\theta_0-\dot p_t}{p_t}\right)\tag{C}. 
\end{equation} \label{condC}
The proof of this theoretical statement is presented in Section \ref{Appendix}. \\

Remark: Condition (C) shows that the time-varying parameter $\theta_t$ becomes unbounded when $p_t = 0$ or $p_t = 1$. Therefore, we consider the following truncated prediction function
\begin{equation}
  p_t^{\epsilon} = \left\{
  \begin{array}{@{}cl@{}}
    \epsilon \!\!\!&\text{if \quad} p_t  < \epsilon  \\
    p_t  \!\!\!&\text{if \quad}  \epsilon \leq p_t < 1 - \epsilon \\
    1 - \epsilon  \!\!\!&\text{if \quad}  p_t \geq 1 - \epsilon
 \end{array}\right.  \label{corrforecast}
\end{equation}
that satisfies $p_t^{\epsilon} \in [\epsilon, 1 - \epsilon]$ for any $0 < \epsilon < \frac{1}{2}$ and $t \in [0,T]$, providing that $\theta_t$ is bounded for every $t \in [0,T]$. \\
From now on, we will keep the notation $p_t$ to denote the truncated prediction function (\ref{corrforecast}), unless specified otherwise.

%{\color{red} The expression for $\theta_t$ is important: $\theta_t$ depens on $t$ through $\dot{p_t}$ and $p_t$. Besides it is parsimonious because only one unknown parameter appear in \ref{theta_t}. Recall it is an upped bound for the real inequality. }

%\subsubsection*{Change of Variables:}
\subsection{A model specification for the forecast error}
 After applying to (\ref{ourmodel}) the simple change of variables $$V_t = X_t - p_t \,,$$ we may introduce the following model for the forecast error of the normalized wind power production:

\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dV_t \!\!\!&=  - \theta_t V_t  \, dt + \sqrt{2 \alpha \theta_0 (V_t +p_t ) (1-V_t-p_t)} \, dW_t\,, \quad t \in [0,T]  \\
   V_0  \!\!\!&=  v_0\in [-1 + \epsilon,1 - \epsilon] \,.
 \end{array}\right.  \label{VtSDE}
\end{equation}

%---BEGIN SECTION 3---
\section{State independent diffusion term: Lamperti transform} \label{Section_3}

Our model (\ref{VtSDE}) for the forecast error has a diffusion term that depends on the state variable $V_t$. To estimate the unknown model parameters, a recommended technique is to modify the SDE (\ref{VtSDE}) by applying the so-called Lamperti transform (see \cite[40--41]{iacus1}, \cite{moma}, \cite[98--100]{saso}) to the process $V$ to obtain a SDE for the transformed process whose diffusion term does not depend anymore on the state variable. To this purpose, we consider the following Lamperti transformation  
%{\color{red} (Remind that continuous-time models with different diffusion term are singular and the likelihood cannot be written down. Besides, it can be unstable to work with low frequency. If the frequency of the data is too low, the Lamperti transform helps to detect this problem (in principle, it is not clear how to determine the right frequency for estimation. )}
%\begin{equation}
%Z_t = h(V_t, t) = \frac{1}{\sqrt{2 \theta_0 \alpha}} \int_{\xi}^{x} \frac{1}{\sqrt{(u + p_t)(1 - u - p_t)}} \,du \Bigg{\vert}_{x = V_t} \,,
%\end{equation}
%where $\xi$ is an arbitrary point of the state space of the process $V$. The choice of $\xi = \frac{1}{2} - p_t$, where $p_t$ is a known input, 
\begin{align}
Z_t = h(V_t, t)  & = \frac{1}{\sqrt{2 \alpha \theta_0 }} \int \frac{1}{\sqrt{(v + p_t)(1 - v - p_t)}} \,dv \Bigg{\vert}_{v = V_t}  \nonumber \\
& = - \sqrt{ \frac{2}{ \alpha \theta_0 }} \arcsin (\sqrt{ 1 - V_t - p_t}) \label{eq:LampZ}
\end{align}
that, after applying It\^{o}'s formula on $h(V_t, t)$, leads to the following SDE with state independent unit diffusion term
\begin{align}
dZ_t = \Bigg[  & \frac{\dot{p}_t}{ \sqrt{2 \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}}  \nonumber \\ 
& + \frac{- \theta_t V_t}{ \sqrt{2  \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}} - \frac{1}{4} \frac{\sqrt{2 \alpha \theta_0} \left( 1 - 2 (V_t + p_t)\right)}{\sqrt{(V_t + p_t)(1 - V_t - p_t)}}  \Bigg] \,dt + dW_t \,. \label{eq:stindepSDE}
\end{align}
After replacing $V_t = 1 - p_t - \sin^2 \left(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \right) $ in (\ref{eq:stindepSDE}), we obtain that the  process $Z$ satisfies the SDE
\begin{align}
dZ_t & = \Bigg[ \frac{ \dot{p}_t  - \theta_t  \left(1 - p_t - \sin^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\sqrt{2 \alpha \theta_0} \cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)}  \nonumber \\
&  - \frac{1}{4}  \frac{\sqrt{2 \alpha \theta_0} \left(1 - 2 \cos^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)} \Bigg] \,dt + dW_t \nonumber \\
&  = \left[  \frac{  2  \dot{p}_t - \theta_t (1 - 2 p_t)  + (\alpha \theta_0 - \theta_t) \cos(- \sqrt{2 \alpha \theta_0 } Z_t) }{\sqrt{2 \alpha \theta_0} \sin{(- \sqrt{2 \alpha \theta_0} Z_t)}}  \right] \,dt + dW_t \,.  \label{eq:stindepSDE2}
\end{align}

{\color{red} We can see in Figure (\ref{fig:LP_transitions}) the effect of the Lamperti transformation upon the forecast error data.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_t_LP.eps}\\
\includegraphics[width=0.45\textwidth]{plots/HP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_t_LP.eps}
\caption{Transitions after the Lamperti transform. The transitions now present a normal Gaussian shape. This effect motivates the use of Gaussian proxies to approximate the process after using the Lamperti transform.}
  \label{fig:LP_transitions}
\end{figure}

The Lamperti transformation has greatly reduced the forecast error skewness, ensuring that the process stays in the range $[0,1]$. Therefore, in this case, the transition densities of the process Z can be adequately approximated through Gaussian densities.}

@@

%---BEGIN SECTION 4---
\section{Likelihood in $V$ space} \label{Section_4}

\subsection{Likelihood}

Suppose that any of $M$ non-overlapping paths of the continuous-time It\^{o} process $V = \{ V_t, t  \in [0,T] \}$ is sampled at $N + 1$ equispaced discrete points with given length interval $\Delta$, and let $ V^{M,N + 1}=\{ V_{t_1^{N + 1}} , V_{t_2^{N + 1}} ,\ldots , V_{t_M^{N + 1}} \}$ denote this random sample, with $V_{t_j^{N + 1}} = \{ V_{t_j + i \Delta}\,, i = 0, \ldots, N \}, \, j = 1, \ldots, M$.

Let $\rho(v \vert v_{j, i-1} ; \bm{\theta})$ be the conditional probability density of $V_{t_j + i \Delta} \equiv V_{j, i}$ given $V_{j, i-1} = v_{j, i-1}$ evaluated at $v$, where $\bm{\theta} = (\theta_0, \alpha, \epsilon)$ are the unknown model parameters.

@@

To guarantee the good behave for our optimization algorithm, we aim to start the optimization as close as we can from the optimal parameters. We use \textbf{Lease Square Minimization} and \textbf{Quadratic Variation} over the data to find an initial guess, and to tune the hyper-parameter $\epsilon\in(0,\frac{1}{2})$.

@@

%---BEGIN SECTION 7---
\section{Results} \label{Section_7}

\subsection{Initial guesses}

>>>>

@@

\setcounter{page}{1}

\begin{document}

@@

Previous attempt by (\cite{mozuma}) considered stochastic wind power forecast models based on stochastic differential equations. Here, we propose an improved model featuring time derivative tracking of the forecast, time-dependent mean reversion, modified diffusion and non-Gaussian approximations. We apply the model to Uruguayan wind power forecasts together with historical wind power production data pertaining to the year 2019. \\

In Section \ref{Section_2}, we introduce the main characteristics of a real data set encompassing the normalized wind power production in Uruguay during the period April-December 2019, joint with the most accurate predictions, as highlighted in our posterior analysis, performed by one out of the three sources of forecast providers. The significant steps for constructing phenomenological models of the normalized wind power production and the forecast error based on stochastic differential equations are described in Section \ref{Section_3}. After reformulating the SDE for the forecast error in Section \ref{Section_4}, by applying the Lamperti transform that leads to a unit diffusion coefficient, we write down in Section \ref{Section_5} the expressions for the likelihood functions of the forecast error in its original space and in the Lamperti space. We also derive simple approximations of the likelihood functions. Section \ref{Section_5} concludes with the description of the optimization algorithm to compute approximate maximum likelihood estimates, including the case where we expand the model comprising an initial transition from the time the forecast is performed to the time of the first forecast.  
In Section \ref{Section_6}, we apply our proposed parametric estimation procedure to the Uruguay wind and forecast dataset, comparing two alternative models and assessing, for the best candidate model, the performance of three different forecast providers.
The proofs of the existence, strong uniqueness, and boundedness of the solutions of the SDE used to model normalized wind power production and its forecast error are given in the Appendix.

@@

%---BEGIN SECTION 2---

\section{Wind power production data in Uruguay and forecast providers } \label{Section_2}

In recent years, Uruguay has triggered a remarkable change in its energy matrix. In (\cite{irena}, p.23) Uruguay is among those countries showcasing innovation, like Denmark, Ireland, Germany, Portugal, and Spain, with proven feasibility of managing annual variable renewable energy (VRE) higher than 25\% in power systems.

According to (\cite{ren21}, pp.118-119), in 2018, Uruguay achieved 36\% of its electricity production from variable wind energy and solar PV, raising the share of generation from wind energy more than five-fold in just four years, from 6.2\% in 2014 to 33\% in 2018.

At present, Uruguay is fostering even higher levels of wind penetration by boosting regional power trading with Argentina and Brazil. 
In this rapidly evolving scenario, it is essential to analyze national data on wind power production joint with wind power short-term forecastings to orientate and assess the strategies and decisions of wind energy actors and businesses.

Our study is based on publicly available data (source: Administrator of Electric Market) on the wind power production in Uruguay for the period April-December 2019, that we adequately normalized with respect to the present $\SI{1474}{\mega\watt}$ maximum installed wind power capacity. Each day, wind power production recordings are available every ten minutes.  In this work, we have also considered data from three different forecast providers, available each day starting at 1 pm.

The next Figure \ref{fig:sample_data} shows the wind power real production during four segments 24-hour long selected from the observation period together with their corresponding hourly short-term forecast, computed by a forecast provider. For the sake of visualization clarity, this Section relies only on forecasts from one provider, called "provider A" from now on, ranked as the most accurate forecast provider, as it emerged from our posterior analysis.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/four_days/377.eps}
\includegraphics[width=0.45\textwidth]{plots/four_days/417.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/four_days/437.eps}
\includegraphics[width=0.45\textwidth]{plots/four_days/477.eps}
\caption{Four 24-hour segments with the wind power real production in Uruguay (orange line) recorded every ten minutes, and the hourly wind power production forecasted by provider A (blue line).}
  \label{fig:sample_data}
\end{figure}

A view of the global discrepancy between the real production and the forecasted production, during the nine months observation period, is summarized through the forecast error histograms, Figure \ref{fig:data_curtailing}, where we also partitioned the forecast errors according to three contiguous categories of normalized generated power. Low normalized generated power corresponds to the range $[0,0.3]$, mid-power refers to the range $(0.3,0.6]$, and high-power to the range $(0.6,1]$.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_6.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_6.eps}
\caption{Wind production forecast error histograms during the period April-December 2019: low-power (upper-left graph), mid-power (upper-right graph), high-power (lower-left graph), and the global range of power (lower-right graph).}
  \label{fig:data_curtailing}
\end{figure}

We may observe that all the histograms in the Figure \ref{fig:data_curtailing} exhibit skewed patterns, to a different extent, as well as extreme observations. The presence of these features can be partly explained. The analysis of the dataset highlighted that, during several 24-hour segments, the system operators decided to reduce, or even cease the wind power production. Indeed, as recalled in (\cite{irena2}, p.3), ``Uruguay experiences high curtailment levels because generation exceeds demand.'' Despite the large country’s interconnection capacity with Argentina and Brazil, there is no active cross-border market, the energy is traded via ad hoc short-term agreements. ``Even with interconnection capacity exceeding peak demand, the power system experiences high VRE curtailment, mostly at night when wind generation exceeds demand.''

The curtailment of the wind power production imposed by the system operators has a strong influence on the forecast error. To build a model that, driven by the available forecast, allows to include the true power production with a prescribed degree of uncertainty, it is necessary to remove the data segments affected by wind curtailment.

Once we removed all the 24-hour segments showing wind curtailment, we set up a dataset containing 147 daily segments. 
In the absence of the curtailment intervention, the forecast error histograms are shown below in Figure \ref{fig:data_after_clean}, where it can be appreciated skewness reduction, except for low power forecast errors histogram.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP.eps}
\includegraphics[width=0.45\textwidth]{plots/MP.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP.eps}
\includegraphics[width=0.45\textwidth]{plots/AP.eps}
\caption{Wind production forecast error histograms during the period April-December 2019 after removing 24-hour segments with artificial wind curtailment: low-power (upper-left graph), mid-power (upper-right graph), high-power (lower-left graph), and the global range of power (lower-right graph).}
  \label{fig:data_after_clean}
\end{figure}

In this stage of data preprocessing, we obtain another useful result by applying the first-order difference operator to the forecast errors. 
The forecast error transition histograms, displayed in the next Figure \ref{fig:error_transitions}, will constitute later a reference for the visual assessment of the global fit of the proposed models.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_t.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_t.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_t.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_t.eps}
\caption{Forecast error transition histograms during the period April-December 2019 without wind power production curtailment: low-power (upper-left graph), mid-power (upper-right graph), high-power (lower-left graph), and the global range of power (lower-right graph).}
\label{fig:error_transitions}
\end{figure}

The histograms in Figure \ref{fig:error_transitions} feature a non-Gaussianity trait and provide initial input for the model-building stage, which is described in Section next.
Later, in Section \ref{Section_4}, guided from inferring the unknown model parameters, we will also propose transforming data as a strategy that naturally leads to an alternative model.

%---BEGIN SECTION 3---

\section{Phenomenological  Model} \label{Section_3}

After analyzing the available dataset, we are now in the condition to build a type of phenomenological model for the normalized wind power generation forecasts that, in its most general form, is a stochastic process $X = \{X_t, \, t \in [0, T] \}$  defined by the following stochastic differential equation (SDE):

\begin{equation}
\begin{cases}
    \dif X_t &=  a(X_t; p_t, \dot{p}_t,\bm{\theta})\dif t + b (X_t; p_t, \dot{p}_t, \bm{\theta} )\dif W_t, \quad t \in [0,T]  \\
     X_0  &=  x_0 \in [0,1],
\end{cases}
 \label{eq:main}
\end{equation}

@@

\subsection{Physical Constraints} \label{Physical_Constraints}

@@

Remark: Observe that a mean-reverting model without derivative tracking shows a delayed path behavior. For instance, consider the diffusion model (\ref{eq:main}) with $a(X_t; p_t, \bm{\theta}) = - \theta_0 (X_t - p_t)\,, \theta_0 > 0$. In this case, given  $ \mathbb{E} \left[X_0\right] = p_0$, the diffusion has mean $\mathbb{E} \left[X_t\right] = p_t - e^{- \theta_0 t } \int_0^t \dot{p}_s  e^{\theta_0 s} \dif s$. The next Figure (\ref{fig:derivative_tracking}) illustrates how different behave the estimated confidence bands for two diffusion models with and without derivative tracking, fitting the same daily segment.

\begin{figure}[H]
\centering
  \includegraphics[width=0.45\textwidth]{plots/31.eps}\includegraphics[width=0.45\textwidth]{plots/31_2.eps}
  \caption{Pointwise confidence bands fitted, for the same daily segment, through diffusion models without derivative tracking (graph on the left) and with derivative tracking (graph on the right).}
  \label{fig:derivative_tracking}
\end{figure}

@@

However, a distinctive feature in our proposed model 
\begin{equation}
\begin{cases}
    \dif X_t&= \left(\dot{p}_t  - \theta_t (X_t - p_t) \right) \dif t +\sqrt{2 \alpha \theta_0 X_t (1 - X_t)} \dif W_t, \quad t \in [0,T]  \\
   X_0 &=  x_0\in [0,1] \,,
\end{cases}
\label{ourmodel}
\end{equation}

is that the drift term contains the time-varying parameter $\theta_t$, rendering the solution $X$ of (\ref{ourmodel}) to a non-stationary and time-inhomogeneous process. To ensure that the process $X$ is the unique strong solution of (\ref{ourmodel}) for all $t \in [0,T]$ with state space $[0,1]$ a.s., the mean-reversion time-varying parameter must satisfy the condition:

\begin{equation}
\theta_t\geq \max\left(\frac{\alpha\theta_0+\dot p_t}{1-p_t},\frac{\alpha\theta_0-\dot p_t}{p_t}\right)\tag{B}. 
\end{equation} \label{condB}
The proof of this theoretical statement is presented in Section \ref{Appendix}. \\

Remark: Condition (B) shows that the time-varying parameter $\theta_t$ becomes unbounded when $p_t = 0$ or $p_t = 1$. Therefore, we consider the following truncated prediction function
\begin{equation}
  p_t^{\epsilon} = \left\{
  \begin{array}{@{}cl@{}}
    \epsilon \!\!\!&\text{if \quad} p_t  < \epsilon  \\
    p_t  \!\!\!&\text{if \quad}  \epsilon \leq p_t < 1 - \epsilon \\
    1 - \epsilon  \!\!\!&\text{if \quad}  p_t \geq 1 - \epsilon
 \end{array}\right.  \label{corrforecast}
\end{equation}
that satisfies $p_t^{\epsilon} \in [\epsilon, 1 - \epsilon]$ for any $0 < \epsilon < \frac{1}{2}$ and $t \in [0,T]$, providing that $\theta_t$ is bounded for every $t \in [0,T]$. \\

Remark: In this work, the analysis of the three different forecast datasets shows that there exists, for any forecast provider, a small $\epsilon >0$ to define the truncated prediction function fulfilling the above condition. \\

From now on, we will keep the notation $p_t$ to denote the truncated prediction function (\ref{corrforecast}), unless specified otherwise.

\subsection{A model specification for the forecast error}
 After applying to (\ref{ourmodel}) the simple change of variables $$V_t = X_t - p_t \,,$$ we may introduce the following model for the forecast error of the normalized wind power production:

\begin{equation}
\begin{cases}
\dif V_t&=  - \theta_t V_t\dif t + \sqrt{2 \alpha \theta_0 (V_t +p_t ) (1-V_t-p_t)}\dif W_t, \quad t \in [0,T]  \\
V_0&=  v_0\in [-1 + p_0,1 - p_0].
\end{cases}  \label{VtSDE}
\end{equation}

@@

%---BEGIN SECTION 4---

\section{State independent diffusion term: Lamperti transform} \label{Section_4}

Our model (\ref{VtSDE}) for the forecast error has a diffusion term that depends on the state variable $V_t$. Under the conditions that permit to use It\^{o}'s formula on a well-chosen transformation of the process $V$, John Lamperti (\cite{lamp}) first showed that the transformed process is again a diffusion that is solution to a SDE with unit coefficient for the diffusion term. 
A vast literature refers nowadays to this result as the so-called Lamperti transform (see, for example, \cite[40--41]{iacus1}, \cite{moma}, \cite[199--203]{pani}, \cite[98--100]{saso}), which is a basic tool to obtain a SDE for the transformed process whose diffusion term does not depend anymore on the state variable. A remarkable effect of removing the state dependency from the random noise term is to increase the numerical stability of the simulated paths of the transformed process. 
For this reason, some estimation methods of the unknown parameters of non-linear SDE models incorporated the use on the Lamperti's change of variable as part of a more complex approximation procedure (for example, the local linearization method in  \cite{shoz}, or the expansion method in \cite{ait}, later extended to multivariate diffusion in \cite{ait2}).

We consider the following Lamperti transformation  
%{\color{red} (Remind that continuous-time models with different diffusion term are singular and the likelihood cannot be written down. Besides, it can be unstable to work with low frequency. If the frequency of the data is too low, the Lamperti transform helps to detect this problem (in principle, it is not clear how to determine the right frequency for estimation. )}
%\begin{equation}
%Z_t = h(V_t, t) = \frac{1}{\sqrt{2 \theta_0 \alpha}} \int_{\xi}^{x} \frac{1}{\sqrt{(u + p_t)(1 - u - p_t)}} \,du \Bigg{\vert}_{x = V_t} \,,
%\end{equation}
%where $\xi$ is an arbitrary point of the state space of the process $V$. The choice of $\xi = \frac{1}{2} - p_t$, where $p_t$ is a known input, 
\begin{align}
Z_t = h(V_t, t; \bm{\theta} )  & = \frac{1}{\sqrt{2 \alpha \theta_0 }} \int \frac{1}{\sqrt{(v + p_t)(1 - v - p_t)}} \dif v \Bigg{\vert}_{v = V_t}   = - \sqrt{ \frac{2}{ \alpha \theta_0 }} \arcsin (\sqrt{ 1 - V_t - p_t}) \label{eq:LampZ}
\end{align}
that, after applying It\^{o}'s formula on $h(V_t, t)$, leads to the following SDE with state independent unit diffusion term
\begin{align}
\dif Z_t = \Bigg[  & \frac{\dot{p}_t}{ \sqrt{2 \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}}  + \frac{- \theta_t V_t}{ \sqrt{2  \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}} - \frac{1}{4} \frac{\sqrt{2 \alpha \theta_0} \left( 1 - 2 (V_t + p_t)\right)}{\sqrt{(V_t + p_t)(1 - V_t - p_t)}}  \Bigg] \dif t + \dif W_t \,. \label{eq:stindepSDE}
\end{align}
After replacing $V_t = 1 - p_t - \sin^2 \left(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \right) $ in (\ref{eq:stindepSDE}), we obtain that the  process $Z$ satisfies the SDE
\begin{align}
\dif Z_t & = \left[ \frac{ \dot{p}_t  - \theta_t  \left(1 - p_t - \sin^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\sqrt{2 \alpha \theta_0} \cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)}   - \frac{1}{4}  \frac{\sqrt{2 \alpha \theta_0} \left(1 - 2 \cos^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)} \right]\dif t + \dif W_t \nonumber \\
&  = \left[  \frac{  2  \dot{p}_t - \theta_t (1 - 2 p_t)  + (\alpha \theta_0 - \theta_t) \cos(- \sqrt{2 \alpha \theta_0 } Z_t) }{\sqrt{2 \alpha \theta_0} \sin{(- \sqrt{2 \alpha \theta_0} Z_t)}}  \right] \dif t + \dif W_t.  \label{eq:stindepSDE2}
\end{align}

We can see in Figure (\ref{fig:LP_transitions}) how the Lamperti transformation modifies the forecast error transition histogram without curtailment displayed in Figure (\ref{fig:error_transitions}).

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_t_LP.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_t_LP.eps}
\caption{Lamperti transformed forecast error transition histograms during the period April-December 2019 without wind power production curtailment: low-power (upper-left graph), mid-power (upper-right graph), high-power (lower-left graph), and the global range of power (lower-right graph).}
  \label{fig:LP_transitions}
\end{figure}

The shape of the forecast error transition histograms after the Lamperti transformation has much resemblance with a Gaussian distribution, motivating toward the use of Gaussian-like approximations of the unknown density transition functions of the process Z.

%{\color{red} Remind that Lamperti transform has the purpose tho check the consistency of the models, that is if the estimation of $\bm{\theta}$ with $V$ and $Z$ differs it is alarming.}

%---BEGIN SECTION 5---

\section{ Likelihood functions of the forecast errors data and optimization algorithm} \label{Section_5}

\subsection{Likelihood in the $V-$space}

Suppose that any of $M$ non-overlapping paths of the continuous-time It\^{o} process $V = \{ V_t, t  \in [0,T] \}$ is sampled at $N + 1$ equispaced discrete points with given length interval $\Delta$, and let $ V^{M,N + 1}=\left\{ V_{t_1^{N + 1}} , V_{t_2^{N + 1}} ,\ldots , V_{t_M^{N + 1}} \right\}$ denote this random sample, with $V_{t_j^{N + 1}} =\left\{ V_{t_j + i \Delta}\,, i = 0, \ldots, N \right\}, \, j = 1, \ldots, M$.

Let $\rho(v \vert v_{j, i-1} ; \bm{\theta})$ be the conditional probability density of $V_{t_j + i \Delta} \equiv V_{j, i}$ given $V_{j, i-1} = v_{j, i-1}$ evaluated at $v$, where $\bm{\theta} = (\theta_0, \alpha)$ are the unknown model parameters.

@@

To guarantee the good behave for our optimization algorithm, we aim to start the optimization as close as we can from the optimal parameters. We use \textbf{Least Square Minimization} and \textbf{Quadratic Variation} over the data to find an initial guess.

@@

%---BEGIN SECTION 7---

\section{Conclusions} \label{Section_7}
